NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-06-12 Manuscript/manuscriptPS.tex

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of calcium binding
are crucial for the PS mediated signaling processes.

>>>>

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of cation binding
are crucial for the PS mediated signaling processes.

Previous experimental studies have concluded that
PS headgroups have more rigid conformation than
phophocholines (PC) due to hydrogen bonding or
electrostatic interactions \cite{browning80,buldt81}.
Multivalent cations or Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids,
while monovalent ions interact more weakly with PS
containing bilayers \cite{hauser77,kurland79,hauser83,hauser85,feigenson86,mattai89,roux90,roux91}.
Dilution of PS rich bilayers with PC lipids
reduces the rigidity and propensity of the
strong complexes with multivalent ions \cite{browning80,buldt81,roux90,roux91}.
